The unequal distribution of access to education and transport: insights from

### 2019-03-18 paper.Rmd

The policy experiment results suggest that a policy that concentrates the provision of public schools spatially is regressive, in the sense that teens living in areas that had the lowest levels of accessibility are the ones who experience the largest negative impact from the policy. They are likely to be worse off compared to an already difficult situation of low local schooling provision, low access by public transport and/or high competition by teens in other areas. In practice, such a change would probably imply larger commuting times for students residing in areas with low accessibility, which can have further consequences in terms of attendance, drop-out likelihood and performance.

Although the simulated scenario represents a rather extreme change in the distribution of public schooling, the way in which students with different levels of accessibility are impacted remains. In the absence of a public transport subsidy for school commutes, students that experience a significant increase in travel-to-school times may be force to change their commute mode from an active (and free of charge mode) to a paid-by mode, with the consequent increase in their monetary spent in commuting to school. The most negative effect is then felt by low-incomes students living in areas with low school accessibility, for whom the marginal increase in commuting costs represents a much larger share in their budget.

@@

Our study can be further extended to account for other modes of school commute, such as active travel. It can also benefit from a finer spatial disaggregation to capture within area inequalities, which may be substantial and are currently hidden away in area aggregates. A future extension of the index can incorporate a measure of school quality, which in so far assumed away. It will also be highly informative to calculate the proposed indices for other contexts and for different time periods in order to understand the true magnitude of the extent of inequality across individuals living in different areas of the SPMR.

We conclude that extending the concept of local accessibility indicators to education can help to both contest and constructively tackle embedded social inequalities. Contestation can be seen as a powerful tool to prevent the implementation of regressive policies that threaten to accentuate already large disparities. Quantitative measures such as the ones proposed here help summarizing multiple dimensions of inequality into a single, interpretable measure. Having this tool may be useful for both policy makers and social groups in discussing and negotiating policies with a highly sensitive social component and evident short and long-term consequences for those affected.

>>>>

The policy experiment results suggest that a policy that concentrates the provision of public schools spatially is regressive. Young people living in areas with the lowest levels of accessibility experienced the largest negative impact from the policy, compounding issues of low local schooling provision, low access by public transport and/or high competition to access good schools. The policy would increase commuting times for students residing in areas with low accessibility. This would likely have further negative knock-on effects in terms of attendance, drop-out likelihood and performance.

Although the simulated scenario represents a rather extreme change in the distribution of public schooling, the way in which students with different levels of accessibility are impacted remains, regardless of the level of concentration. In the absence of a public transport subsidy for school commutes, students who experience a substantial increase in travel-to-school times may be forced to change their commute mode from an active (and free of charge mode) to a paid-by mode, with the consequent increase in their monetary spent in commuting to school. The most negative effect is then felt by low-incomes students living in areas with low school accessibility, for whom the marginal increase in commuting costs represents a much larger share in their budget.

@@

Our study could be further extended to account for other modes of school commute, such as active travel. It would benefit from data with higher spatial resolution to capture within-area inequalities, which may be substantial but currently hidden away in area aggregates. A future extension of the index could also incorporate measures of school quality. Furthermore, it would be informative to calculate the proposed indices for other contexts and different time periods, to understand the extent to which the regressive nature of school agglomeration is a universal finding or specific to the SPMR.

We conclude that extending the concept of local accessibility indicators to education can help to both contest and constructively tackle embedded social inequalities. Contestation can be seen as a powerful tool to prevent the implementation of regressive policies that threaten to accentuate already large disparities. Quantitative measures such as those proposed here can help demonstrate how multiple dimensions of inequality amplify inequalities, via a single interpretable measure. The methods demonstrated could be useful for policy makers and campaigners to negotiate policies with a highly sensitive social component and evident short and long-term consequences for those affected.
